How environmental drivers of synchrony interact

### 2022-01-17 Paper/Paper.Rnw
thorough edit of theoretical case study results now complete

%***DAN: Right now sup mat section S7 is not cited. It only has figures and tables, not sure I plan to 
%add anything to it other than that. Either get rid of it as a section, or else cite it in some of the places 
%where you currently cite the figures and tables in it. My current hunch is that the first of those options is
%better.

@@

%***DAN: stopped here

%Case study A, new try
CaA helps provide an intuitive understanding about how lags in Moran drivers can produce contrasting interactions between
Moran effects on different timescales, and how our theory captures that contrast.
For CaA, $\epsilon^{(1)}$ effects on populations were lagged by $1$q but 
$\epsilon^{(2)}$ effects were unlagged (Methods). So, in the language of Fig. \ref{fig:IntroTheory},
$l_{e1}=1$q and $l_{e2}=0$q. 
The between-noise lag, $l_n$, of Fig. \ref{fig:IntroTheory} is $0$q because $(\epsilon^{(1)},\epsilon^{(2)})$ was 
taken to be a white noise process (see SI section \ref{SM-sect:examples_cross_sync} for details, here).
Thus, what determines the sign of interactions between Moran effects for this example is how 
$l_{e1}-l_{e2}+l_n=1$q compares to the timescale/period, $\sigma$, being considered. 
On the shortest timescales ($\sigma=0.5 \text{y} = 2 \text{q}$; Fig. \ref{fig:TheoryCaseA}a, left side of panel), $1$q
is half the period, so interactions effects are negative. On long timescales (e.g., $\sigma>8$y), $1$q is a negligible portion of the
period, so $l_{e1}-l_{e2}+l_n=1$q is close to zero, relative to that period, and interaction effects
are positive: compared to such long timescales, $\epsilon^{(1)}$ and $\epsilon^{(2)}$ effects happen close to simultaneously, so
the two noise variables reinforce each other. Theory captures these complexities because
the size of the $1$q lag of this scenario, relative to a timescale, $\sigma$, is captured by
the phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}}$, which equals the phase of
$f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$ because $l_n=0$ and $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ is a real number
(SI section \ref{SM-sect:examples_cross_sync}). But $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}=
p_1^{(1)} p_0^{(2)} \mu = p_1^{(1)} p_0^{(2)} \exp(-2 \pi \iota f) = 
p_1^{(1)} p_0^{(2)} \exp(-2 \pi \iota /\sigma)$ (SI section \ref{SM-sect:details_case_studies}), 
which has phase $-\pi$ at short timescales, $\sigma \approx 2 \text{q}$,
and phase almost $0$ at long timescales (Fig. \ref{fig:TheoryCaseA}b).
The phase controls the sign of $2\Rp(f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}})$, the
interaction term in (\ref{rhoww_T1to3}), as described in General Theory. Comparing
Fig. \ref{fig:TheoryCaseA}a,d shows the additional influence of autoregressive population effects, which
simply multiply all Moran influences by the same timescale-dependent non-negative 
quantity, not altering their relative importance or sign [see (\ref{rhoww_T1to3})].

@@

%Case study B, new version
CaB reveals how Moran effects can interact when lags are longer than 1 model time step. 
For CaB, $\epsilon^{(1)}$ effects on populations were lagged by $n$q but 
$\epsilon^{(2)}$ effects were unlagged (Methods). We used $n=3$, so $l_{e1}=3$q and $l_{e2}=0$q. 
As for CaA, because $(\epsilon^{(1)},\epsilon^{(2)})$ was 
a white noise process for CaB, $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ was a real number and $l_n=0$.
Interaction effects were again negative on short timescales 
($\sigma=0.5 \text{y} = 2 \text{q}$; Fig. \ref{fig:TheoryCaseA}e, far left side of panel)
because $l_{e1}-l_{e2}+l_n=3$q was $1.5\sigma$ on that timescale, and so effects of 
$\epsilon^{(1)}$ and $\epsilon^{(2)}$ were in a half-phase relationship and
counteracted each other.
Similar to CaA, $l_{e1}-l_{e2}+l_n=3$q was again negligible compared to long timescales, so interaction effects
were positive on long timescales (Fig. \ref{fig:TheoryCaseA}e, right side of panel).
The quantity $l_{e1}-l_{e2}+l_n=3$q exactly equaled the timescale for $\sigma=3 \text{q} = 0.75 \text{y}$ and 
equaled half the timescale for $\sigma=6 \text{q} = 1.5 \text{y}$, hence interactions effects
were positive for $\sigma=0.75 \text{y}$ and negative for $\sigma=1.5$y 
(Fig. \ref{fig:TheoryCaseA}e). Theory captures these complexities because 
the size of the $3$q relative lag of this scenario, relative to $\sigma$,
is again captured by the phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$ (Fig. \ref{fig:TheoryCaseA}f).

@@

%***DAN: stopped here

@@

%Case study C, new
CaA and CaB assumed white noise processes, and therefore real-valued $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
and $l_n=0$ (SI sections \ref{SM-sect:examples_cross_sync}).
But CaC illustrates what can happen when noise processes have periodic components, e.g., 
both processes oscillate with annual periodicity, an important scenario because seasonality is common.
For CaC, $\epsilon^{(1)}$ effects on populations were lagged by $n_1$q and 
$\epsilon^{(2)}$ effects were lagged by $n_2$q (Methods). We used $n_1=1$, $n_2=0$, $p_1^{(1)}=3$ and $p_0^{(2)}=-1.5$, so 
$\epsilon^{(1)}$ effects were lagged and positive, and $\epsilon^{(2)}$ effects were immediate and negative
(Fig. \ref{fig:TheoryCaseA} caption). The lagged positive effects of $\epsilon^{(1)}$ correspond to 
$l_{e1}=1$q, in the language of Fig. \ref{fig:IntroTheory}. But, for that figure, we assumed environmental 
variables had positive effects on populations, whereas for CaC $\epsilon^{(2)}$ had negative effects.
Immediate negative effects of an annually oscillating variable are the same as positive effects lagged 
by $2$q, so we can take $l_{e2}=2$q for CaC.  
%***DAN: pretty confused, come back to this

%Case study C, old
CaA and CaB assumed white noise processes, and therefore real-valued $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
(SI sections \ref{SM-sect:examples_cross_sync}, \ref{SM-sect:details_case_studies}),
but CaC illustrates what can happen when noise processes have periodic components, e.g., 
both processes oscillate with annual periodicity, an important scenario because seasonality is common. 
What matters then is whether and how the lag between the two seasonal noise processes aligns with the difference 
between the lags of their effects on populations, as we now explain. 
For the scenarios we considered under CaC (Fig. \ref{fig:TheoryCaseA}i-p), 
$\epsilon^{(1)}$ effects on populations were positive, and were 
lagged (i.e., delayed in their influence) by 1 quarter; but 
$\epsilon^{(2)}$ effects were immediate and negative (Methods and Fig. \ref{fig:TheoryCaseA} caption).
For CaC1 (Fig. \ref{fig:TheoryCaseA}i-l), peaks in the periodic
noise process $\epsilon^{(1)}$ lagged peaks in the periodic process $\epsilon^{(2)}$ by
1 quarter. Thus lags compounded: the annual positive population effects of $\epsilon^{(1)}$ 
were exactly 2 quarters offset from the annual negative effects of $\epsilon^{(2)}$, 
producing reinforcing interactions 
between Moran effects on annual timescales (Fig. \ref{fig:TheoryCaseA}i), and more population
synchrony on annual timescales than would have been expected if Moran effects operated independently. Contrastingly,
for CaC2 (Fig. \ref{fig:TheoryCaseA}m-p), peaks in the periodic
noise process $\epsilon^{(2)}$ lagged peaks in the periodic process $\epsilon^{(1)}$ by
1 quarter. Thus lags cancelled: the annual positive population effects of $\epsilon^{(1)}$
coincided with the negative effects of $\epsilon^{(2)}$ every year, a half-phase shift for an 
annually oscillating effect, which corresponds to negative interactions between Moran influences
and less population synchrony, on annual timescales, than would have been expected if Moran effects operated independently 
(Fig. \ref{fig:TheoryCaseA}m). The only difference between CaC1 and CaC2 
is the relative timing of annual peaks in $\epsilon^{(1)}$ and $\epsilon^{(2)}$.
Our theory captures these complexities
because $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$
quantifies the relative phase lag between the effects of $\epsilon^{(1)}$ and 
$\epsilon^{(2)}$ compared to a given timescale, while 
$\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ captures the phase lag between 
oscillations in the variables themselves, compared to the timescale (Fig. \ref{fig:TheoryCaseA}j,k,n,o).
The phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}}$
therefore captures whether relative lags of effects and lags between 
the variables themselves align or misalign, with the degree of alignment
reflected in the real part of this quantity, $\Rp(f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}})$,
which is half the interaction term in our theory (\ref{rhoww_T1to3}).
On long timescales, interactions were the
same in the two scenarios because, on those timescales, sub-annual lags 
make negligible difference. Interactions were then negative 
because the effects of $\epsilon^{(1)}$ on populations were 
positive and those of $\epsilon^{(2)}$ were negative.

>>>>

%***DAN: Right now sup mat section S7 is not cited. It only has figures and tables, not sure I plan to 
%add anything to it other than that. Either get rid of it as a section, or else cite it in some of the places 
%where you currently cite the figures and tables in it. My current hunch is that the first of those options is
%better. Have to decide what to do with Fig. S1 in that case.

@@

%Case study A, new try
CaA helps provide an intuitive understanding about how lags in Moran drivers can produce contrasting interactions between
Moran effects on different timescales, and how our theory captures that contrast.
For CaA, $\epsilon^{(1)}$ effects on populations were lagged by $1$q but 
$\epsilon^{(2)}$ effects were unlagged (Methods). So, in the language of Fig. \ref{fig:IntroTheory},
$l_{e1}=1$q and $l_{e2}=0$q. 
The between-noise lag, $l_n$, of Fig. \ref{fig:IntroTheory} is $0$q because $(\epsilon^{(1)},\epsilon^{(2)})$ was 
taken to be a white noise process (see SI section \ref{SM-sect:examples_cross_sync} for details, here).
Thus, what determines the sign of interactions between Moran effects for this example is how 
$l_{e1}-l_{e2}+l_n=1$q compares to the timescale/period, $\sigma$, being considered. 
On the shortest timescales ($\sigma=0.5 \text{y} = 2 \text{q}$; Fig. \ref{fig:TheoryCaseA}a, left side of panel), $1$q
is half the period, so interactions effects are negative. On long timescales (e.g., $\sigma>8$y), $1$q is a negligible portion of the
period, so $l_{e1}-l_{e2}+l_n=1$q is close to zero, relative to that period, and interaction effects
are positive: compared to such long timescales, $\epsilon^{(1)}$ and $\epsilon^{(2)}$ effects happen close to simultaneously, so
the two noise variables reinforce each other. Theory captures these complexities because
the size of the $1$q lag of this scenario, relative to a timescale, $\sigma$, is captured by
the phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}}$, which equals the phase of
$f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$ because $l_n=0$ and $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ is a real number
(SI section \ref{SM-sect:examples_cross_sync}). But $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}=
p_1^{(1)} p_0^{(2)} \mu = p_1^{(1)} p_0^{(2)} \exp(-2 \pi \iota f) = 
p_1^{(1)} p_0^{(2)} \exp(-2 \pi \iota /\sigma)$ (SI section \ref{SM-sect:details_case_studies}), 
which has phase $-\pi$ at short timescales, $\sigma \approx 2 \text{q}$,
and phase almost $0$ at long timescales (Fig. \ref{fig:TheoryCaseA}b,c).
The phase controls the sign of $2\Rp(f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}})$, the
interaction term in (\ref{rhoww_T1to3}), as described in General Theory. Comparing
Fig. \ref{fig:TheoryCaseA}a,d shows the additional influence of autoregressive population effects, which
simply multiply all Moran influences by the same timescale-dependent non-negative 
quantity, not altering their relative importance or sign [see (\ref{rhoww_T1to3})].

@@

%Case study B, new version
CaB reveals how Moran effects can interact when lags are longer than 1 model time step. 
For CaB, $\epsilon^{(1)}$ effects on populations were lagged by $n$q but 
$\epsilon^{(2)}$ effects were unlagged (Methods). We used $n=3$, so $l_{e1}=3$q and $l_{e2}=0$q. 
As for CaA, because $(\epsilon^{(1)},\epsilon^{(2)})$ was 
a white noise process for CaB, $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ was a real number and $l_n=0$.
Interaction effects were again negative on short timescales 
($\sigma=0.5 \text{y} = 2 \text{q}$; Fig. \ref{fig:TheoryCaseA}e, far left side of panel)
because $l_{e1}-l_{e2}+l_n=3$q was $1.5\sigma$ on that timescale, and so effects of 
$\epsilon^{(1)}$ and $\epsilon^{(2)}$ were in a half-phase relationship and
counteracted each other.
Similar to CaA, $l_{e1}-l_{e2}+l_n=3$q was again negligible compared to long timescales, so interaction effects
were positive on long timescales (Fig. \ref{fig:TheoryCaseA}e, right side of panel).
The quantity $l_{e1}-l_{e2}+l_n=3$q exactly equaled the timescale for $\sigma=3 \text{q} = 0.75 \text{y}$ and 
equaled half the timescale for $\sigma=6 \text{q} = 1.5 \text{y}$, hence interactions effects
were positive for $\sigma=0.75 \text{y}$ and negative for $\sigma=1.5$y 
(Fig. \ref{fig:TheoryCaseA}e). Theory captures these complexities because 
the size of the $3$q lag of this scenario, relative to $\sigma$,
is again captured by the phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$ (Fig. \ref{fig:TheoryCaseA}f,g).

@@

%Case study C, new
CaA and CaB assumed white noise processes, and therefore real-valued $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
and $l_n=0$.
But CaC illustrates what can happen when noise processes have periodic components and lagged interdependences, e.g., 
both processes oscillate with annual periodicity and distinct phenology, an important scenario because seasonality is common.
For CaC, $\epsilon^{(1)}$ effects on populations were lagged by $n_1$q and 
$\epsilon^{(2)}$ effects were lagged by $n_2$q (Methods). We used $n_1=1$, $n_2=0$, $p_1^{(1)}=3$ and $p_0^{(2)}=-1.5$, so 
$\epsilon^{(1)}$ effects were lagged and positive, and $\epsilon^{(2)}$ effects were immediate and negative. The language and logic of 
Fig. \ref{fig:IntroTheory} does not directly apply, because we there assumed, in order to introduce concepts simply,
that environmental effects were positive, and we now take $p_0^{(2)}=-1.5$. But similar logic does apply, with
both the current and Fig. \ref{fig:IntroTheory} paradigms covered by our general theory.
For CaC1 (Fig. \ref{fig:TheoryCaseA}i-l), peaks in the periodic
noise process $\epsilon^{(1)}$ lagged peaks in the periodic process $\epsilon^{(2)}$ by
$1$q (Methods, SI section \ref{SM-sect:examples_cross_sync}). Thus lags compounded: the annual positive population effects of $\epsilon^{(1)}$ 
were exactly $2$q offset from the annual negative effects of $\epsilon^{(2)}$, 
producing reinforcing interactions 
between Moran effects on annual timescales (Fig. \ref{fig:TheoryCaseA}i).
Contrastingly, for CaC2 (Fig. \ref{fig:TheoryCaseA}m-p), peaks in the periodic
noise process $\epsilon^{(2)}$ lagged peaks in the periodic process $\epsilon^{(1)}$ by
$1$q. Thus lags cancelled: the annual positive population effects of $\epsilon^{(1)}$
coincided with the negative effects of $\epsilon^{(2)}$ every year, producing 
negative interactions between Moran influences (Fig. \ref{fig:TheoryCaseA}m).
Our theory captures these complexities
because $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$
quantifies the degree of alignment between the effects of $\epsilon^{(1)}$ and 
$\epsilon^{(2)}$ compared to a given timescale, while 
$\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ captures the degree of aslignment between 
oscillations in the variables themselves, compared to the timescale (Fig. \ref{fig:TheoryCaseA}j,k,n,o).
The phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}}$
therefore captures whether delays in effects and lags between 
variables align or misalign, with alignment
reflected in the real part of this quantity, $\Rp(f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}})$,
which is half the interaction term in our theory (\ref{rhoww_T1to3}).
On long timescales, interactions were the
same in the two scenarios, and were negative, because, on those timescales, sub-annual lags 
make negligible difference, and the effects of $\epsilon^{(1)}$ on populations were 
positive and those of $\epsilon^{(2)}$ were negative.

% %Case study C, old
% CaA and CaB assumed white noise processes, and therefore real-valued $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$
% (SI sections \ref{SM-sect:examples_cross_sync}, \ref{SM-sect:details_case_studies}),
% but CaC illustrates what can happen when noise processes have periodic components, e.g., 
% both processes oscillate with annual periodicity, an important scenario because seasonality is common. 
% What matters then is whether and how the lag between the two seasonal noise processes aligns with the difference 
% between the lags of their effects on populations, as we now explain. 
% For the scenarios we considered under CaC (Fig. \ref{fig:TheoryCaseA}i-p), 
% $\epsilon^{(1)}$ effects on populations were positive, and were 
% lagged (i.e., delayed in their influence) by 1 quarter; but 
% $\epsilon^{(2)}$ effects were immediate and negative (Methods and Fig. \ref{fig:TheoryCaseA} caption).
% For CaC1 (Fig. \ref{fig:TheoryCaseA}i-l), peaks in the periodic
% noise process $\epsilon^{(1)}$ lagged peaks in the periodic process $\epsilon^{(2)}$ by
% 1 quarter. Thus lags compounded: the annual positive population effects of $\epsilon^{(1)}$ 
% were exactly 2 quarters offset from the annual negative effects of $\epsilon^{(2)}$, 
% producing reinforcing interactions 
% between Moran effects on annual timescales (Fig. \ref{fig:TheoryCaseA}i), and more population
% synchrony on annual timescales than would have been expected if Moran effects operated independently. Contrastingly,
% for CaC2 (Fig. \ref{fig:TheoryCaseA}m-p), peaks in the periodic
% noise process $\epsilon^{(2)}$ lagged peaks in the periodic process $\epsilon^{(1)}$ by
% 1 quarter. Thus lags cancelled: the annual positive population effects of $\epsilon^{(1)}$
% coincided with the negative effects of $\epsilon^{(2)}$ every year, a half-phase shift for an 
% annually oscillating effect, which corresponds to negative interactions between Moran influences
% and less population synchrony, on annual timescales, than would have been expected if Moran effects operated independently 
% (Fig. \ref{fig:TheoryCaseA}m). The only difference between CaC1 and CaC2 
% is the relative timing of annual peaks in $\epsilon^{(1)}$ and $\epsilon^{(2)}$.
% Our theory captures these complexities
% because $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}}$
% quantifies the relative phase lag between the effects of $\epsilon^{(1)}$ and 
% $\epsilon^{(2)}$ compared to a given timescale, while 
% $\rho_{\epsilon^{(1)} \epsilon^{(2)}}$ captures the phase lag between 
% oscillations in the variables themselves, compared to the timescale (Fig. \ref{fig:TheoryCaseA}j,k,n,o).
% The phase of $f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}}$
% therefore captures whether relative lags of effects and lags between 
% the variables themselves align or misalign, with the degree of alignment
% reflected in the real part of this quantity, $\Rp(f_{\mcP^{(1)}} \conj{f_{\mcP^{(2)}}} \rho_{\epsilon^{(1)} \epsilon^{(2)}})$,
% which is half the interaction term in our theory (\ref{rhoww_T1to3}).
% On long timescales, interactions were the
% same in the two scenarios because, on those timescales, sub-annual lags 
% make negligible difference. Interactions were then negative 
% because the effects of $\epsilon^{(1)}$ on populations were 
% positive and those of $\epsilon^{(2)}$ were negative.

@@
